---
title: "Bi610_Assignment4b"
author: "Carina Ahrens"
date: "2024-11-27"
output: html_document
---
# Fully Loaded Lizards

```{r}
library("lme4")
library("Matrix")
library("ggplot2")

themePlot <- 
  theme(plot.title = element_text(hjust = 0.5)) +
  theme_bw() +
  theme(plot.title = element_text(color = "black",  margin = NULL, lineheight = NULL, size = 20, hjust = .5, vjust = .5, face = "plain")) +
  theme(legend.text = element_text(color = "black", margin = NULL, lineheight = NULL, size = 10, hjust = .5, vjust = .5, face = "plain")) +
  theme(legend.title = element_text(color = "black",  margin = NULL, lineheight = NULL, size = 20, hjust = .5, vjust = .5, face = "plain")) +
  theme(legend.key.height = unit(0.3, 'cm'), legend.key.size = unit(0.5,'cm'), legend.key.width = unit(0.5,'cm')) +
  theme(axis.text.x = element_text(color = "black", size = 20, hjust = .5, vjust = .5, face = "plain")) +
  theme(axis.title.x = element_text(color = "black", size = 20, hjust = .5, vjust = .5, face = "plain")) +
  theme(axis.text.y = element_text(color = "black", size = 20, hjust = .5, vjust = .5, face = "plain")) +
  theme(axis.title.y = element_text(color = "black", size = 20, hjust = .5, vjust = .5, face = "plain"))
```

```{r read in data}
lizards <- read.table("FullyLoadedLizards.tsv", sep=' ', header=TRUE, quote='"')
lizards$Species <- lizards$species
lizards[lizards$Species %in% c("bearded dragon"), "Species"] <-  "Dragon"
lizards[lizards$Species %in% c("chameleon"), "Species"] <-  "Chameleon"
lizards[lizards$Species %in% c("crested gecko"), "Species"] <-  "Gecko"
```


```{r explore data}
names(lizards)
table(lizards$Species)
table(lizards$lighting)
table(lizards$tank)
table(lizards$substrate)
table(lizards$individuals)#? not sure what this is.. individuals per category?

# they seem to be evenly distributed over the categories
table(lizards$species, lizards$lighting)
table(lizards$species, lizards$tank)
table(lizards$species, lizards$substrate)
```

### Variables:
* species
  * bearded dragons (360)
  * chameleons (360)
  * crested geckos (360)
* lighting:
  * heat (270)
  * heat + UVA  (270)
  * heat + UVB  (270)
  * heat + UVA & UVB  (270)
* tank:
  * 10-gallon (360)
  * 20-gallon (360)
  * 40-gallon (360)
* substrate
  * natural (540)
  * unnatural (540)
* individuals:
* weight (in grams):
* mean_weight:

RQ: which conditions are optimal for growth for each species and if there is any interaction between the different living conditions

Focus in Report: please report your findings of optimum growing conditions for each species (which tank size, light, and substrate shows the greatest growth over the measurement period) and any interactions, positive or negative, you find between any of the different living conditions

```{r creating subsets}
dragon <- subset(lizards, subset=species %in% c("bearded dragon"))
chameleon <- subset(lizards, subset=species %in% c("chameleon"))
gecko <- subset(lizards, subset=species %in% c("crested gecko"))
```


```{r visualization - weight all lizards}
par(mfrow = c(1, 1))
hist(lizards$weight, xlab="Weight", breaks = 100,  main="Weight - All Lizards")
abline(v=mean(lizards$weight), col='red', lwd=2)
abline(v=mean(dragon$weight), col='blue', lwd=1)
abline(v=mean(chameleon$weight), col='green', lwd=1)
abline(v=mean(gecko$weight), col='orange', lwd=1)
```

```{r visualization - mean weight all lizards}
par(mfrow = c(1, 1))
hist(lizards$mean_weight, xlab="Weight", breaks = 100,  main="Weight - All Lizards")
abline(v=mean(lizards$mean_weight), col='red', lwd=1)
abline(v=mean(lizards$mean_weight), col='red', lwd=1)
# not sure what mean_weight is
```


```{r visualization - weight by species}
par(mfrow = c(1, 3))
hist(dragon$weight, xlab="Weight", breaks = 100,  main="Weight - Bearded Dragon")
abline(v=mean(dragon$weight), col='red', lwd=1)
hist(chameleon$weight, xlab="Weight", breaks = 100,  main="Weight - Chameleon")
abline(v=mean(chameleon$weight), col='red', lwd=1)
hist(gecko$weight, xlab="Weight", breaks = 100,  main="Weight - Crested Gecko")
abline(v=mean(gecko$weight), col='red', lwd=1)
```
# ```{r}
# library(ggplot2)
# ggplot(lizards, aes(x = weight, y = lighting, colour = species)) +
#   geom_point()+
#   #scale_x_reverse() + scale_y_reverse() +
#   scale_color_manual(values = c("red", "darkblue", "green")) +
#   ggtitle("FACE T3 40%")
#   #mytheme +
#   #ggsave("16q T3_40.jpg", width = 30, height = 20, units = "cm", dpi = 600)
# ```

```{r, interaction plot}
# interaction plot - strain x-axis & response y-axis different colors for different genotypes
ggplot(lizards, aes(x = species, y = weight, group = lighting, color = lighting)) +
  stat_summary(fun = "mean", geom = "point", size = 1.5) +
  stat_summary(fun = "mean", geom = "line") +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = .2) +
  scale_color_manual(values = c("blue", "red", "green4", "orange")) +
  ggtitle("Weight by Species and Lighting") +
  labs(x="Species", y="Weight", colour = "Lighting") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.title = element_text(color = "black",  margin = NULL, lineheight = NULL, size = 25, hjust = .5, vjust = .5, face = "plain")) +
  theme(legend.text = element_text(color = "black", margin = NULL, lineheight = NULL, size = 15, hjust = .5, vjust = .5, face = "plain")) +
  theme(legend.title = element_text(color = "black",  margin = NULL, lineheight = NULL, size = 20, hjust = .5, vjust = .5, face = "plain")) +
  theme(legend.key.height = unit(0.3, 'cm'), legend.key.size = unit(0.5,'cm'), legend.key.width = unit(0.5,'cm')) +
  theme(axis.text.x = element_text(angle = 90, color = "black", size = 15, hjust = .5, vjust = .5, face = "plain")) +
  theme(axis.title.x = element_text(color = "black", size = 20, hjust = .5, vjust = .5, face = "plain")) +
  theme(axis.text.y = element_text(color = "black", size = 15, hjust = .5, vjust = .5, face = "plain")) +
  theme(axis.title.y = element_text(color = "black", size = 20, hjust = .5, vjust = .5, face = "plain"))
```

```{r}
VPlot_species0 <- 
ggplot(lizards, aes(x = factor(species), y = weight, fill = species)) +
  geom_violin(scale = "width", trim = FALSE, alpha = 0.7) +
  scale_fill_manual(values = c("blue", "green4", "red")) +
  labs(x = "Species", y = "Weight", fill = "Species") +
  themePlot
VPlot_species0
# main effect of species - bearded dragons = heigher weight
```

```{r}
VPlot_lighting <- 
ggplot(lizards, aes(x = factor(lighting), y = weight, fill = species)) +
  geom_violin(scale = "width", trim = FALSE, alpha = 0.7) +
  scale_fill_manual(values = c("blue", "green4", "red")) +
  labs(x = "Lighting", y = "Weight", fill = "Species") +
  themePlot
VPlot_lighting
# effect of light
```
```{r}
VPlot_tank <- 
ggplot(lizards, aes(x = factor(tank), y = weight, fill = species)) +
  geom_violin(scale = "width", trim = FALSE, alpha = 0.7) +
  scale_fill_manual(values = c("blue", "green4", "red")) +
  labs(x = "Tank", y = "Weight", fill = "Species") +
  themePlot
VPlot_tank
# maybe interaction species & tank (sig. for bearded dragon)
```

```{r}
VPlot_substrate <- 
ggplot(lizards, aes(x = factor(substrate), y = weight, fill = species)) +
  geom_violin(scale = "width", trim = FALSE, alpha = 0.7) +
  scale_fill_manual(values = c("blue", "green4", "red")) +
  labs(x = "Substrate", y = "Weight", fill = "Species") +
  themePlot
VPlot_substrate
#no effect substrate 
```

```{r}
VPlot_lighting2 <- 
ggplot(lizards, aes(x = factor(species), y = weight, fill = lighting)) +
  geom_violin(scale = "width", trim = FALSE, alpha = 0.7) +
  scale_fill_manual(values = c("blue", "green4", "red", "orange")) +
  labs(x = "Species", y = "Weight", fill = "Lighting") +
  themePlot
VPlot_lighting2
```
```{r}
VPlot_tank2 <- 
ggplot(lizards, aes(x = factor(species), y = weight, fill = tank)) +
  geom_violin(scale = "width", trim = FALSE, alpha = 0.7) +
  scale_fill_manual(values = c("blue", "green4", "red", "orange")) +
  labs(x = "Species", y = "Weight", fill = "Tank") +
  themePlot
VPlot_tank2
# probably main effect of tank
```
```{r}
VPlot_substrate2 <- 
ggplot(lizards, aes(x = factor(species), y = weight, fill = substrate)) +
  geom_violin(scale = "width", trim = FALSE, alpha = 0.7) +
  scale_fill_manual(values = c("blue", "green4", "red", "orange")) +
  labs(x = "Species", y = "Weight", fill = "Substrate") +
  themePlot
VPlot_substrate2
```

```{r}
summary(lizards_M01 <- lmer(weight ~ 1 + species  + lighting + tank + substrate + (1|individuals), lizards))
drop1(lizards_M01, test="Chisq")
summary(lizards_M02 <- update(lizards_M01, .~. - substrate), correlation = FALSE)
drop1(lizards_M02, test="Chisq")
```

```{r}
summary(lizards_M01 <- lmer(weight ~ 1 + species + species:lighting + species:tank + substrate + (1|individuals), lizards))
drop1(lizards_M01, test="Chisq")
summary(lizards_M02 <- update(lizards_M01, .~. - substrate), correlation = FALSE)
drop1(lizards_M02, test="Chisq")
```

```{r}
summary(lizards_M01_SUM <- lmer(weight ~ 1 + species + species:lighting + species:tank + substrate + (1|individuals), lizards,  contrasts = list(species = "contr.sum", lighting = "contr.sum", tank = "contr.sum")))
drop1(lizards_M01_SUM, test="Chisq")
summary(lizards_M02_SUM <- update(lizards_M01, .~. - substrate), correlation = FALSE)
drop1(lizards_M02_SUM, test="Chisq")
```