---
title: "BI610_Assignment4"
author: "Leah Blankenship, James Maze, Carina Ahrens"
date: "2024-11-21"
output: html_document
---


## Humanities (Carina)

This data set is set up to show an impact of h-index for humanities. Graduates from the humanities with a higher h-index tend to have a higher annual income.
Therefore the data is simulated in two subsets:
The first subset contains 700 out of 1000 sampled academics whose papers show a lower impact in the field of humanities (lower h-index). The second subset contains the remaining 300 data points with higher h-indexes (= higher impact) and thus a higher salary.
The data sets are generated using a normal distribution and estimates for mean salary and h-indexes. Additionally, information of completing a postdoc and location of where the Ph.D. was obtained are randomly assigned in a way that they will not show to have an impact on the annual salary for graduates from humanities.

```{r for reproducablity}
library(dplyr)
library(ggplot2)

set.seed(12)
```

```{r humanities_1 - parameters}
# creating parameters for humanity population (lower salary group)
# like pumpkin example
humanities_grads_salary_params <- list(
  mu = 90000, # mean yearly income (low income group in humanities)
  alpha = c('n' = -5000, # post doc position (yes vs no)
            'y' = +20000),
  beta = c('us' = 0, # location of Ph.D. (U.S. vs international)
           'int' = +1000
  ),
  # No interaction
  sigma = 20000
)
```

```{r humanities_1 - distribution}
# like pumpkin example
# creating salary for population with parameters
distribution <- rnorm(100000, humanities_grads_salary_params$mu, humanities_grads_salary_params$sigma) # normally distributed -> rnorm

sample_size <- 700 # only sampled part of data (part of lower income group in humanities)

humanities_grads <- expand.grid( #setting up data frame with columns for postDoc, location and salary
  post_doc = rep(NA, sample_size), 
  location = rep(NA),
  salary = NA
)

#for-loop to fill data frame with information about postDoc and location (set up to not have an impact on annual salary)
for (r in 1:nrow(humanities_grads)) {
  # randomly assign postDoc or not with 50/50 proportion
  humanities_grads$post_doc[r] <- sample(c('y', 'n'), size=1, prob=c(0.5, 0.5))
  # randomly assign location  with 50/50 proportion
  humanities_grads$location[r] <- sample(c('us', 'int'), size=1, prob=c(0.5, 0.5))
}
```

```{r humanities_1 - PhD & location}
# assigning PhD & location impact here
for (j in 1:nrow(humanities_grads)) {
  pd <- as.character(humanities_grads$post_doc[j])
  loc <- as.character(humanities_grads$location[j])       
  gamma <- 0 # no interaction wanted here
  humanities_grads$salary[j] <- (
    sample(distribution, 1)                      
    + humanities_grads_salary_params$alpha[pd]  
    + humanities_grads_salary_params$beta[loc]                
    + gamma                          
  )
}
```

```{r humanities_1 - histogram distribution salary}
# hist(humanities_grads$salary)
# abline(v=mean(humanities_grads$salary), col='red', lwd=1)
```

```{r humanities_1 h-index}
#adding impact of h-index (mean ~10 -> normal/low impact on salary)  - trial and error to get a reasonable number here
humanities_grads <- humanities_grads |> 
  dplyr::mutate(h_index = humanities_grads$salary * 0.00005 + rnorm(700, 10, 4))
mean_h_index <- mean(humanities_grads$h_index) # 14.97711 seems realistic
correlation_h_salary <- cor(humanities_grads$salary, humanities_grads$h_index) # 0.3209226
range_h_index <- range(humanities_grads$h_index) # ~1.8 - ~30.92
```

```{r humanities_1 - distribution h-index}
# hist(humanities_grads$h_index)
# abline(v=mean_h_index, col='red', lwd=1)
```

```{r humanities_2 - parameters}
# Now follows the creation of data of the group in humanities with a higher h-index and a higher salary.
# like pumpkin example
# creating parameters for humanity population (higher salary group)
humanities_grads_H_salary_params <- list(
  mu = 140000, # mean yearly income -> set to be higher then for the first group
  alpha = c('n' = -5000,
            'y' = +20000
  ),
  beta = c('us' = 0,
           'int' = +1000
  ),
  # No interaction
  sigma = 20000
)
```

```{r humanities_2 - distribution}
# like pumpkin example
distribution <- rnorm(100000, humanities_grads_H_salary_params$mu, humanities_grads_H_salary_params$sigma) # also normal distribution

sample_size <- 300 # rest of population to reach 1000 data points for humanities

#setting up data frame with columns for postDoc, location and salary
humanities_grads_H <- expand.grid(
  post_doc = rep(NA, sample_size), 
  location = rep(NA),
  salary = NA
)

#for-loop to fill data frame with information about postdoc and location (set up to not have an effect)
for (r in 1:nrow(humanities_grads_H)) {
  # randomly assign post-doc or not with 50/50 proportion
  humanities_grads_H$post_doc[r] <- sample(c('y', 'n'), size=1, prob=c(0.5, 0.5))
  # randomly assign location with proportion (0.4 international, 0.6 U.S.)
  humanities_grads_H$location[r] <- sample(c('us', 'int'), size=1, prob=c(0.5, 0.5))
}
```

```{r humanities_2 - PhD & location}
for (j in 1:nrow(humanities_grads_H)) {
  pd <- as.character(humanities_grads_H$post_doc[j])
  loc <- as.character(humanities_grads_H$location[j])        
  gamma <- 0  # no interaction wanted
  humanities_grads_H$salary[j] <- (
    sample(distribution, 1)
    + humanities_grads_H_salary_params$alpha[pd]
    + humanities_grads_H_salary_params$beta[loc]                
    + gamma                          
  )
}
```

```{r humanities_2 - histogram distribution salary}
# hist(humanities_grads_H$salary)
# abline(v=mean(humanities_grads_H$salary), col='red', lwd=1)
```

```{r humanities_2 - h-index}
#adding impact of h-index (mean ~25 -> higher impact on salary) - trial and error to get a reasonable number here
humanities_grads_H <- humanities_grads_H |> 
  dplyr::mutate(h_index = humanities_grads_H$salary * 0.00005 + rnorm(300, 25, 5))
mean_h_index_H <- mean(humanities_grads_H$h_index) # 32.57963 seems realistic
correlation_h_salary_H <- cor(humanities_grads_H$salary, humanities_grads_H$h_index) # 0.343683
range_h_index_H <- range(humanities_grads_H$h_index) # ~14.06 ~48
```

```{r humanities_2 histogram distribution h-index}
# hist(humanities_grads_H$h_index)
# abline(v=mean(humanities_grads_H$h_index), col='red', lwd=1)
```

```{r humanities - adding together data frames of low and high income group}
humanities <- rbind(humanities_grads, humanities_grads_H)
humanities$field <- "humanities" # adding column for field, so data can be combined later
```

```{r humanities - histogram salary distribution}
# histogram for full humanities salary
# hist(humanities$salary) # a little screwed; as it was set up!
# abline(v=mean(humanities_grads$salary), col='red', lwd=1)
# abline(v=mean(humanities_grads_H$salary), col='red', lwd=1)
# abline(v=mean(humanities$salary), col='blue', lwd=1)
```

```{r humanities - histogram h-index distribution}
# histogram for full humanities salary
# hist(humanities$h_index) # a little screwed to the right; as it was set up!
# abline(v=mean_h_index, col='red', lwd=1)
# abline(v=mean_h_index_H, col='red', lwd=1)
# abline(v=mean(humanities$h_index), col='blue', lwd=1)
```

```{r humanities - scatter plot distribution salary vs h-index}
# plot(
#   x = humanities$salary,
#   y = humanities$h_index,
#   cex = 1,
#   pch = 21,
# )
```

## Simulating Data for Physics and Chemistry (Carina)
For graduates of Physics and Chemistry the data set is set up to not show an impact of h-index, location or post doctorate after graduating but graduates of Physics and Chemistry on average have a higher annual salary (e.g. in comparison to graduates from the humanities).

```{r PhyChem - parameters}
# creating parameters for humanity population (higher salary than humanities)
# like pumpkin example
Phy_Chem_grads_salary_params <- list(
  mu = 150000, # mean yearly income
  alpha = c('n' = -5000, # post doc position (yes vs no)
            'y' = +20000
  ),
  beta = c('us' = 0, # location of Ph.D. (U.S. vs international)
           'int' = +1000
  ),
  # No interaction
  sigma = 20000
)
```

```{r PhyChem - distribution}
# creating salary for population with parameters
distribution <- rnorm(100000, Phy_Chem_grads_salary_params$mu, Phy_Chem_grads_salary_params$sigma) 

sample_size <- 1000

Phy_Chem_grads <- expand.grid( #setting up data frame with columns for postDoc, location and salary
  post_doc = rep(NA, sample_size), 
  location = rep(NA),
  salary = NA
)

for (r in 1:nrow(Phy_Chem_grads)) {
  # randomly assign postDoc or not with 50/50 proportion
  Phy_Chem_grads$post_doc[r] <- sample(c('y', 'n'), size=1, prob=c(0.5, 0.5))
  # randomly assign postDoc or not with 50/50 proportion
  Phy_Chem_grads$location[r] <- sample(c('us', 'int'), size=1, prob=c(0.5, 0.5))
}
```

```{r PhyChem - PhD & location}
# assigning PhD & location impact here
for (j in 1:nrow(Phy_Chem_grads)) {
  pd <- as.character(Phy_Chem_grads$post_doc[j])
  loc <- as.character(Phy_Chem_grads$location[j])         
  gamma <- 0                       #no interaction wanted
  Phy_Chem_grads$salary[j] <- (
    sample(distribution, 1)
    + Phy_Chem_grads_salary_params$alpha[pd]
    + Phy_Chem_grads_salary_params$beta[loc]                
    + gamma                          
  )
}
```

```{r PhyChem - histogram distribution salary}
 # hist(Phy_Chem_grads$salary)
 # abline(v=mean(Phy_Chem_grads$salary), col='red', lwd=1)
```

```{r PhyChem h-index}
#adding impact of h-index (mean ~20 -> higher impact on salary) - trial and error to get a reasonable number here
Phy_Chem_grads <- Phy_Chem_grads |> 
  dplyr::mutate(h_index = Phy_Chem_grads$salary * 0.00005 + rnorm(1000, 20, 7))
mean_impact <- mean(Phy_Chem_grads$h_index) # 27.5773
correlation <- cor(Phy_Chem_grads$salary, Phy_Chem_grads$h_index) # 0.148731
range_h_index <- range(humanities_grads$h_index) # ~1.8 - ~31
```

```{r PhyChem - histogram h-index distribution}
# hist(Phy_Chem_grads$h_index)
# abline(v=mean_impact, col='red', lwd=1)
```

```{r PhyChem - scatter plot distribution salary vs h-index}
# plot(
#   x = Phy_Chem_grads$salary,
#   y = Phy_Chem_grads$h_index,
#   cex = 1,
#   pch = 21,
# )
```

```{r Phy_Chem - adding field column}
Phy_Chem_grads$field <- "Physics_Chemistry"
```

## Engingeering and Computer Science (Leah)

In order to generate data for this field, I first set parameters for what I would expect the mean salary to be for engineer and computer science graduates based on a Google search. Then I chose a standard deviation and how much that I though post doc status and location of PhD will effect salary in our hypothetical data. 

I used these parameters to generate salary data for 1000 subjects from a normal distribution. Then I used a mutate function with some scaling to generate h-index data that had a low-moderate correlation with salary and an h-index mean that was similar to the value I found through a Google search. 

```{r set parameters, include=FALSE}

eng_comp_salary_params <- list(
  mu = 123000, # based on google search for mean salary
  alpha = c('n' = -5000, # set effect of having a post doc on salary
            'y' = +20000
  ),
  beta = c('us' = 0, #set effect of getting a PhD from the US or internationally on salary
           'int' = 0
  ),
  # Not adding any interaction terms (gamma)
  sigma = 20000 #set standard deviation of population parameters
)

#make population distribution. I chose to make the engineer/computer science salary normally distributed
population_dist <- rnorm(100000, eng_comp_salary_params$mu, eng_comp_salary_params$sigma)

```

```{r generate data, include=FALSE}
sample_size <- 1000

eng_comp_grads <- expand.grid(
  post_doc = rep(NA, sample_size), 
  location = NA,
  salary = NA
)

for (r in 1:nrow(eng_comp_grads)) {
  # randomly assign post-doc or not with proportion (0.3 post-doc, 0.7 not)
  eng_comp_grads$post_doc[r] <- sample(c('y', 'n'), size=1, prob=c(0.3, 0.7))
  # randomly assign location with proportion (0.4 international, 0.6 U.S.)
  eng_comp_grads$location[r] <- sample(c('us', 'int'), size=1, prob=c(0.4, 0.6))
  
}
# generate salary data
for (j in 1:nrow(eng_comp_grads)) {
  pd <- as.character(eng_comp_grads$post_doc[j]) # temp. store post doc status
  loc <- as.character(eng_comp_grads$location[j])      # temp. store location         
  gamma <- 0                       #no interaction, so no combination factor needed
  eng_comp_grads$salary[j] <- (
    sample(population_dist, 1)                        #sample randomly from the population distribution
    + eng_comp_salary_params$alpha[pd]                # assign expected impact of other variables
    + eng_comp_salary_params$beta[loc]                
    + gamma                          
  )
}

#hist(eng_comp_grads$salary) #plot salary histogram

#mutate salary, manually altering scaling variables and mean of index distribution until desired mean impact and desired correlation between impact and salary is reached. Here, I picked mean h-index ~ 11 (based on a google search) and correlation ~0.3 (because we wanted h-index to be only weakly important in this field)

eng_comp_grads <- eng_comp_grads |> 
  dplyr::mutate(h_index = eng_comp_grads$salary * 0.00005 + rnorm(1000, 5, 3))
mean_impact <- mean(eng_comp_grads$h_index)
correlation <- cor(eng_comp_grads$salary, eng_comp_grads$h_index)
#hist(eng_comp_grads$h_index) #plot h-index values

eng_comp_grads$field <- "engineering/comp sci"

```

## Applied Life Sciences (Leah)

To generate this data, I repeated the same process that I used for the engineering and computer science data, but altered the parameters to reflect more realistic numbers for this field where possible. 

```{r set parameters, include=FALSE}
#Repeat the same steps for Applied Life Sciences, with adjustments to salary, h-index means, and effects of post doc achievement
appliedlife_salary_params <- list(
  mu = 100000,
  alpha = c('n' = -4000,
            'y' = +25000
  ),
  beta = c('us' = 0,
           'int' = 0
  ),
  # Not adding any interaction terms (gamma)
  sigma = 15000
)
  
population_dist <- rnorm(100000, appliedlife_salary_params$mu, appliedlife_salary_params$sigma)

```

```{r generate data, include=FALSE}
sample_size <- 1000

appliedlife_grads <- expand.grid(
  post_doc = rep(NA, sample_size), 
  location = rep(NA),
  salary = NA
)

for (r in 1:nrow(appliedlife_grads)) {
  # randomly assign post-doc or not with proportion (0.3 post-doc, 0.7 not)
  appliedlife_grads$post_doc[r] <- sample(c('y', 'n'), size=1, prob=c(0.3, 0.7))
  # randomly assign location with proportion (0.4 international, 0.6 U.S.)
  appliedlife_grads$location[r] <- sample(c('us', 'int'), size=1, prob=c(0.4, 0.6))
  
  # Assign params based on pumpkin example
}

for (j in 1:nrow(appliedlife_grads)) {
  pd <- as.character(appliedlife_grads$post_doc[j]) # temp. store post doc status
  loc <- as.character(appliedlife_grads$location[j])      # temp. store location         
  gamma <- 0                       #no interaction, so no combination factor needed
  appliedlife_grads$salary[j] <- (
    sample(population_dist, 1)                        #sample randomly from the population distribution
    + appliedlife_salary_params$alpha[pd]                # assign expected impact of other variables
    + appliedlife_salary_params$beta[loc]                
    + gamma                          
  )
}

#hist(appliedlife_grads$salary) #plot salary histogram


appliedlife_grads <- appliedlife_grads |> 
  dplyr::mutate(h_index = appliedlife_grads$salary * 0.00001 + rnorm(1000, 1.5, 0.5))
mean_impact <- mean(appliedlife_grads$h_index)
correlation <- cor(appliedlife_grads$salary, appliedlife_grads$h_index)
hist(appliedlife_grads$h_index) #plot h-index values

appliedlife_grads$field <- "applied life sciences"
```

## R Earth Sciences & Biology Distribution (James)

To generate this data, I used a gamma distribution. I made it so that different proportions of respondents conducted post-docs (30% yes, 70% no) and international (40% international, 60% U.S.) PhDs. To match simulated influence on salary, I adjusted the gamma distribution's mean based on the location parameter (U.S. or International) and the post-doc parameter. I calculated the rate parameter as rate = shape / mean, with the shape parameter constant for all Earth Science and Biology Grads (shape=5). For each graduate, I simulated a gamma distribution with 1000 salary observations given the appropriate parameters. Then, I randomly sampled each graduate's gamma distribution to simulate their salary. The h-index was drawn from a normal distribution, and the value was multiplied by salary. By adjusting the standard deviation of the distribution, we can strengthen or weaken the correlation with salary. 

Here's the major findings on Earth Science and Bioloy PhDs:
- Salary follows a gamma distribution with rightward skew
- Conducting a post-doc substantially increases private-sector salary
- International PhDs do not have a discernible impact on salary, given this sample size (power)
- The h-index has a fairly strong correlation to graduate's salary, and h-indices are higher than Math & Stats graduates. 

```{r}
params <- list(
  mu = 95000,
  alpha = c('n' = -9000,
            'y' = +17000
  ),
  beta = c('us' = 0,
           'int' = +1000
  )
)
```

```{r}
sample_size <- 890
gamma_size <- 1000

earth_bio_grads <- data.frame(
  post_doc = rep(NA, sample_size), 
  location = rep(NA, sample_size),
  field = 'Earth Science & Biology',
  salary = NA,
  h_index = NA
)

grads <- nrow(earth_bio_grads)

for (r in 1:grads) {
  # randomly assign post-doc or not with proportion (0.3 post-doc, 0.7 not)
  earth_bio_grads$post_doc[r] <- sample(c('y', 'n'), size=1, prob=c(0.3, 0.7))
  # randomly assign location with proportion (0.4 international, 0.6 U.S.)
  earth_bio_grads$location[r] <- sample(c('us', 'int'), size=1, prob=c(0.4, 0.6))
  
  # Assign params 
  pd <- as.character(earth_bio_grads$post_doc[r])
  loc <- as.character(earth_bio_grads$location[r])
  full_mu = params$mu
  alpha = params$alpha[pd]
  beta = params$beta[loc]
  
  sample_mu <- full_mu + alpha + beta
  
  sample_shape <- 5
  sample_rate <- (sample_shape / sample_mu)
  
  gamma_dist <- rgamma(gamma_size, shape = sample_shape, rate = sample_rate)
  
  # if (r %in% c(5, 99, 900)) {
  #   hist(gamma_dist, breaks = 50, main = paste("Gamma Distribution (Iteration: with", pd, " and", loc,")"),
  #        xlab = "Value", probability = TRUE, col = "lightgreen")
  #   lines(density(gamma_dist), col = "blue", lwd = 2)
  # }
  
  salary <- sample(gamma_dist, size=1)
  
  earth_bio_grads$salary[r] <- salary
}
```

```{r}
variability_param <- 0.5 # Make it bigger to decrease h-index correlation
sd <- variability_param / 2

earth_bio_grads <- earth_bio_grads %>% 
  mutate(h_index = salary * (1 + rnorm(grads, 0, sd))) %>% 
  mutate(h_index = case_when(h_index < 0 ~ 0,
                             h_index >= 0 ~ h_index)) %>% 
  mutate(h_index = h_index / 11000) 

print(cor(earth_bio_grads$h_index, earth_bio_grads$salary))

options(scipen = 999 )

# plot <- ggplot(data = earth_bio_grads, 
#                mapping = aes(x=h_index, 
#                              y=salary, 
#                              color=post_doc)) +
#   geom_point() +
#   #scale_x_continuous(labels = label_number()) +
#   theme_bw() +
#   labs(title="Correlation between h-index and income")
# 
# print(plot)
```

## Math and Statistics Grads (James)

The Math and Statistics graduates were simulated with the same steps as Earth Sciences and Biology graduates; however, the parameters and thus, the overall findings are different. 

- Overall, the mean salary is much higher for Math and Statistics PhDs, and the variance is much higher. (i.e. the gamma distribution has a smaller shape parameter (3.5 instead of 5). This means these graduates will have some extremely high earners. 
- Conducting a post-doc will have a small, but likely un-observable impact on private sector salary. 
- U.S. students Math and Statistics Students generally have higher salaries than international students.
- h-index has a relatively weak correlation with 

```{r}

params <- list(
  mu = 152000,
  alpha = c('n' = -1000,
            'y' = +4000
  ),
  beta = c('us' = +20000,
           'int' = -2500
  )
)
```

```{r}

sample_size <- 525
gamma_size <- 1000

math_stats_grads <- data.frame(
  post_doc = rep(NA, sample_size), 
  location = rep(NA, sample_size),
  field = 'Math & Statistics',
  salary = NA,
  h_index = NA
)

grads <- nrow(math_stats_grads)

for (r in 1:grads) {
  # randomly assign post-doc or not with proportion (0.3 post-doc, 0.7 not)
  math_stats_grads$post_doc[r] <- sample(c('y', 'n'), size=1, prob=c(0.15, 0.85))
  # randomly assign location with proportion (0.4 international, 0.6 U.S.)
  math_stats_grads$location[r] <- sample(c('us', 'int'), size=1, prob=c(0.75, 0.25))
  
  # Assign params 
  pd <- as.character(math_stats_grads$post_doc[r])
  loc <- as.character(math_stats_grads$location[r])
  full_mu = params$mu
  alpha = params$alpha[pd]
  beta = params$beta[loc]
  
  sample_mu <- full_mu + alpha + beta
  
  sample_shape <- 3.5
  sample_rate <- (sample_shape / sample_mu)
  
  gamma_dist <- rgamma(gamma_size, shape = sample_shape, rate = sample_rate)
  
  # if (r %in% c(5, 99, 500)) {
  #   hist(gamma_dist, breaks = 50, main = paste("Gamma Distribution (Iteration: with", pd, " and", loc,")"),
  #        xlab = "Value", probability = TRUE, col = "lightgreen")
  #   lines(density(gamma_dist), col = "blue", lwd = 2)
  # }
  
  salary <- sample(gamma_dist, size=1)
  
  math_stats_grads$salary[r] <- salary
}
```

```{r}

variability_param <- 3 # Make it bigger to decrease h-index correlation
sd <- variability_param / 2


earth_bio_grads <- earth_bio_grads %>% 
  mutate(h_index = salary * (1 + rnorm(grads, 0, sd))) %>% 
  mutate(h_index = case_when(h_index < 0 ~ 0,
                             h_index >= 0 ~ h_index)) %>% 
  mutate(h_index = h_index / 28000) 

print(cor(math_stats_grads$h_index, math_stats_grads$salary))

options(scipen = 999 )

# plot <- ggplot(data = math_stats_grads, 
#                mapping = aes(x=h_index, 
#                              y=salary, 
#                              color=location)) +
#   geom_point() +
#   #scale_x_continuous(labels = label_number()) +
#   theme_bw() +
#   labs(title="Correlation between h-index and income")
# 
# print(plot)

```

```{r combining data frames & exporting}
academia <- rbind(humanities, Phy_Chem_grads, eng_comp_grads, earth_bio_grads, math_stats_grads)
write.csv(academia, file='Bi610_dataSet_academia.csv', row.names=FALSE)
```

```{r academia - checkin data}
#table(academia$field)
#table(academia$post_doc)
#table(academia$location)
```

```{r academia - histogram salary distribution}
# hist(academia$salary)
```

```{r academia - histogram h-index distribution}
# hist(academia$h_index)
```

```{r academia - scatter plot distribution salary vs h-index}
# plot(
#   x = academia$salary,
#   y = academia$h_index,
#   cex = 1,
#   pch = 21,
# )
```
