---
title: "JM_data_simulation"
author: "James Maze"
date: "2024-11-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(scales)
set.seed(42) # If you know, you know ! :)

```

## R Earth Sciences & Biology Distribution (James)

To generate this data, I used a gamma distribution. I made it so that different proportions of respondents conducted post-docs (30% yes, 70% no) and international (40% international, 60% U.S.) PhDs. To match simulated influence on salary, I adjusted the gamma distribution's mean based on the location parameter (U.S. or International) and the post-doc parameter. I calculated the rate parameter as rate = shape / mean, with the shape parameter constant for all Earth Science and Biology Grads (shape=5). For each graduate, I simulated a gamma distribution with 1000 salary observations given the appropriate parameters. Then, I randomly sampled each graduate's gamma distribution to simulate their salary. The h-index was drawn from a normal distribution, and the value was multiplied by salary. By adjusting the standard deviation of the distribution, we can strengthen or weaken the correlation with salary. 

Here's the major findings on Earth Science and Bioloy PhDs:
- Salary follows a gamma distribution with rightward skew
- Conducting a post-doc substantially increases private-sector salary
- International PhDs do not have a discernible impact on salary, given this sample size (power)
- The h-index has a fairly strong correlation to graduate's salary, and h-indices are higher than Math & Stats gradautes. 

```{r}

params <- list(
  mu = 95000,
  alpha = c('n' = -9000,
            'y' = +17000
  ),
  beta = c('us' = 0,
           'int' = +1000
  )
)

```

```{r}
sample_size <- 890
gamma_size <- 1000

earth_bio_grads <- data.frame(
  post_doc = rep(NA, sample_size), 
  location = rep(NA, sample_size),
  field = 'Earth Science & Biology',
  salary = NA,
  h_index = NA
)

grads <- nrow(earth_bio_grads)

for (r in 1:grads) {
  # randomly assign post-doc or not with proportion (0.3 post-doc, 0.7 not)
  earth_bio_grads$post_doc[r] <- sample(c('y', 'n'), size=1, prob=c(0.3, 0.7))
  # randomly assign location with proportion (0.4 international, 0.6 U.S.)
  earth_bio_grads$location[r] <- sample(c('us', 'int'), size=1, prob=c(0.4, 0.6))
  
  # Assign params 
  pd <- as.character(earth_bio_grads$post_doc[r])
  loc <- as.character(earth_bio_grads$location[r])
  full_mu = params$mu
  alpha = params$alpha[pd]
  beta = params$beta[loc]
  
  sample_mu <- full_mu + alpha + beta
  
  sample_shape <- 5
  sample_rate <- (sample_shape / sample_mu)
  
  gamma_dist <- rgamma(gamma_size, shape = sample_shape, rate = sample_rate)
  
  if (r %in% c(5, 99, 900)) {
    hist(gamma_dist, breaks = 50, main = paste("Gamma Distribution (Iteration: with", pd, " and", loc,")"),
         xlab = "Value", probability = TRUE, col = "lightgreen")
    lines(density(gamma_dist), col = "blue", lwd = 2)
  }
  
  salary <- sample(gamma_dist, size=1)
  
  earth_bio_grads$salary[r] <- salary
}

```

```{r}

variability_param <- 0.5 # Make it bigger to decrease h-index correlation
sd <- variability_param / 2

earth_bio_grads <- earth_bio_grads %>% 
  mutate(h_index = salary * (1 + rnorm(grads, 0, sd))) %>% 
  mutate(h_index = case_when(h_index < 0 ~ 0,
                             h_index >= 0 ~ h_index)) %>% 
  mutate(h_index = h_index / 11000) 

print(cor(earth_bio_grads$h_index, earth_bio_grads$salary))

options(scipen = 999 )

plot <- ggplot(data = earth_bio_grads, 
               mapping = aes(x=h_index, 
                             y=salary, 
                             color=post_doc)) +
  geom_point() +
  scale_x_continuous(labels = label_number()) +
  theme_bw() +
  labs(title="Correlation between h-index and income")

print(plot)
```

## Math and Statistics Grads (James)

The Math and Statistics graduates were simulated with the same steps as Earth Sciences and Biology graduates; however, the parameters and thus, the overall findings are different. 

- Overall, the mean salary is much higher for Math and Statistics PhDs, and the variance is much higher. (i.e. the gamma distribution has a smaller shape parameter (3.5 instead of 5). This means these graduates will have some extremely high earners. 
- Conducting a post-doc will have a small, but likely un-observable impact on private sector salary. 
- U.S. students Math and Statistics Students generally have higher salaries than international students.
- h-index has a relatively weak correlation with 



```{r}

params <- list(
  mu = 152000,
  alpha = c('n' = -1000,
            'y' = +4000
  ),
  beta = c('us' = +20000,
           'int' = -2500
  )
)
```

```{r}

sample_size <- 525
gamma_size <- 1000

math_stats_grads <- data.frame(
  post_doc = rep(NA, sample_size), 
  location = rep(NA, sample_size),
  field = 'Math & Statistics',
  salary = NA,
  h_index = NA
)

grads <- nrow(math_stats_grads)

for (r in 1:grads) {
  # randomly assign post-doc or not with proportion (0.3 post-doc, 0.7 not)
  math_stats_grads$post_doc[r] <- sample(c('y', 'n'), size=1, prob=c(0.15, 0.85))
  # randomly assign location with proportion (0.4 international, 0.6 U.S.)
  math_stats_grads$location[r] <- sample(c('us', 'int'), size=1, prob=c(0.75, 0.25))
  
  # Assign params 
  pd <- as.character(math_stats_grads$post_doc[r])
  loc <- as.character(math_stats_grads$location[r])
  full_mu = params$mu
  alpha = params$alpha[pd]
  beta = params$beta[loc]
  
  sample_mu <- full_mu + alpha + beta
  
  sample_shape <- 3.5
  sample_rate <- (sample_shape / sample_mu)
  
  gamma_dist <- rgamma(gamma_size, shape = sample_shape, rate = sample_rate)
  
  if (r %in% c(5, 99, 500)) {
    hist(gamma_dist, breaks = 50, main = paste("Gamma Distribution (Iteration: with", pd, " and", loc,")"),
         xlab = "Value", probability = TRUE, col = "lightgreen")
    lines(density(gamma_dist), col = "blue", lwd = 2)
  }
  
  salary <- sample(gamma_dist, size=1)
  
  math_stats_grads$salary[r] <- salary
}
```

```{r}

variability_param <- 3 # Make it bigger to decrease h-index correlation
sd <- variability_param / 2


earth_bio_grads <- earth_bio_grads %>% 
  mutate(h_index = salary * (1 + rnorm(grads, 0, sd))) %>% 
  mutate(h_index = case_when(h_index < 0 ~ 0,
                             h_index >= 0 ~ h_index)) %>% 
  mutate(h_index = h_index / 28000) 

print(cor(math_stats_grads$h_index, math_stats_grads$salary))

options(scipen = 999 )

plot <- ggplot(data = math_stats_grads, 
               mapping = aes(x=h_index, 
                             y=salary, 
                             color=location)) +
  geom_point() +
  scale_x_continuous(labels = label_number()) +
  theme_bw() +
  labs(title="Correlation between h-index and income")

print(plot)

```



