---
title: "Biostats_Assigment4"
author: "LeahB"
date: "2024-11-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
set.seed(42)

```

## Engingeering and Computer Science (Leah)

In order to generate data for this field, I first set parameters for what I would expect the mean salary to be for engineer and computer science graduates based on a Google search. Then I chose a standard deviation and how much that I though post doc status and location of PhD will effect salary in our hypothetical data. 

I used these parameters to generate salary data for 1000 subjects from a normal distribution. Then I used a mutate function with some scaling to generate h-index data that had a low-moderate correlation with salary and an h-index mean that was similar to the value I found through a Google search. 

```{r set parameters, include=FALSE}

eng_comp_salary_params <- list(
  mu = 123000, # based on google search for mean salary
  alpha = c('n' = -5000, # set effect of having a post doc on salary
            'y' = +20000
  ),
  beta = c('us' = 0, #set effect of getting a PhD from the US or internationally on salary
           'int' = 0
  ),
  # Not adding any interaction terms (gamma)
  sigma = 20000 #set standard deviation of population parameters
)

#make population distribution. I chose to make the engineer/computer science salary normally distributed
population_dist <- rnorm(100000, eng_comp_salary_params$mu, eng_comp_salary_params$sigma)

```

```{r generate data, include=FALSE}
sample_size <- 1000

eng_comp_grads <- expand.grid(
  post_doc = rep(NA, sample_size), 
  location = NA,
  salary = NA
)

for (r in 1:nrow(eng_comp_grads)) {
  # randomly assign post-doc or not with proportion (0.3 post-doc, 0.7 not)
  eng_comp_grads$post_doc[r] <- sample(c('y', 'n'), size=1, prob=c(0.3, 0.7))
  # randomly assign location with proportion (0.4 international, 0.6 U.S.)
  eng_comp_grads$location[r] <- sample(c('us', 'int'), size=1, prob=c(0.4, 0.6))
  
}
# generate salary data
for (j in 1:nrow(eng_comp_grads)) {
  pd <- as.character(eng_comp_grads$post_doc[j]) # temp. store post doc status
  loc <- as.character(eng_comp_grads$location[j])      # temp. store location         
  gamma <- 0                       #no interaction, so no combination factor needed
  eng_comp_grads$salary[j] <- (
    sample(population_dist, 1)                        #sample randomly from the population distribution
    + eng_comp_salary_params$alpha[pd]                # assign expected impact of other variables
    + eng_comp_salary_params$beta[loc]                
    + gamma                          
  )
}

#hist(eng_comp_grads$salary) #plot salary histogram

#mutate salary, manually altering scaling variables and mean of index distribution until desired mean impact and desired correlation between impact and salary is reached. Here, I picked mean h-index ~ 11 (based on a google search) and correlation ~0.3 (because we wanted h-index to be only weakly important in this field)

eng_comp_grads <- eng_comp_grads |> 
  dplyr::mutate(h_index = eng_comp_grads$salary * 0.00005 + rnorm(1000, 5, 3))
mean_impact <- mean(eng_comp_grads$h_index)
correlation <- cor(eng_comp_grads$salary, eng_comp_grads$h_index)
#hist(eng_comp_grads$h_index) #plot h-index values

eng_comp_grads$field <- "engineering/comp sci"

```


## Applied Life Sciences (Leah)

To generate this data, I repeated the same process that I used for the engineering and computer science data, but altered the parameters to reflect more realistic numbers for this field where possible. 
```{r set parameters, include=FALSE}
#Repeat the same steps for Applied Life Sciences, with adjustments to salary, h-index means, and effects of post doc achievement
appliedlife_salary_params <- list(
  mu = 100000,
  alpha = c('n' = -4000,
            'y' = +25000
  ),
  beta = c('us' = 0,
           'int' = 0
  ),
  # Not adding any interaction terms (gamma)
  sigma = 15000
)
  
population_dist <- rnorm(100000, appliedlife_salary_params$mu, appliedlife_salary_params$sigma)

```

```{r generate data, include=FALSE}
sample_size <- 1000

appliedlife_grads <- expand.grid(
  post_doc = rep(NA, sample_size), 
  location = rep(NA),
  salary = NA
)

for (r in 1:nrow(appliedlife_grads)) {
  # randomly assign post-doc or not with proportion (0.3 post-doc, 0.7 not)
  appliedlife_grads$post_doc[r] <- sample(c('y', 'n'), size=1, prob=c(0.3, 0.7))
  # randomly assign location with proportion (0.4 international, 0.6 U.S.)
  appliedlife_grads$location[r] <- sample(c('us', 'int'), size=1, prob=c(0.4, 0.6))
  
  # Assign params based on pumpkin example
}

for (j in 1:nrow(appliedlife_grads)) {
  pd <- as.character(appliedlife_grads$post_doc[j]) # temp. store post doc status
  loc <- as.character(appliedlife_grads$location[j])      # temp. store location         
  gamma <- 0                       #no interaction, so no combination factor needed
  appliedlife_grads$salary[j] <- (
    sample(population_dist, 1)                        #sample randomly from the population distribution
    + appliedlife_salary_params$alpha[pd]                # assign expected impact of other variables
    + appliedlife_salary_params$beta[loc]                
    + gamma                          
  )
}

#hist(appliedlife_grads$salary) #plot salary histogram


appliedlife_grads <- appliedlife_grads |> 
  dplyr::mutate(h_index = appliedlife_grads$salary * 0.00001 + rnorm(1000, 1.5, 0.5))
mean_impact <- mean(appliedlife_grads$h_index)
correlation <- cor(appliedlife_grads$salary, appliedlife_grads$h_index)
hist(appliedlife_grads$h_index) #plot h-index values

#appliedlife_grads$field <- "applied life sciences"
```
  # Not adding any interaction terms (gamma)
  sigma = 20000
)
  
population_dist <- rnorm(100000, eng_comp_salary_params$mu, eng_comp_salary_params$sigma) #Currently using a normal distribution for simplicity. We can change later or use different distributions for other fields

```

```{r}
sample_size <- 1000

eng_comp_grads <- expand.grid(
  post_doc = rep(NA, sample_size), 
  location = rep(NA),
  salary = NA
)

for (r in 1:nrow(eng_comp_grads)) {
  # randomly assign post-doc or not with proportion (0.3 post-doc, 0.7 not)
  eng_comp_grads$post_doc[r] <- sample(c('y', 'n'), size=1, prob=c(0.3, 0.7))
  # randomly assign location with proportion (0.4 international, 0.6 U.S.)
  eng_comp_grads$location[r] <- sample(c('us', 'int'), size=1, prob=c(0.4, 0.6))
  
  # Assign params based on pumpkin example
}

for (j in 1:nrow(eng_comp_grads)) {
  pd <- as.character(eng_comp_grads$post_doc[j]) # temp. store post doc status
  loc <- as.character(eng_comp_grads$location[j])      # temp. store location         
  gamma <- 0                       #no interaction, so no combination factor needed
  eng_comp_grads$salary[j] <- (
    sample(population_dist, 1)                        #sample randomly from the population distribution
    + eng_comp_salary_params$alpha[pd]                # assign expected impact of other variables
    + eng_comp_salary_params$beta[loc]                
    + gamma                          
  )
}

hist(eng_comp_grads$salary) #plot salary histogram

```


```{r}
#mutate salary, altering scaling variables and mean of index distibution until desired mean impact and desired correlation between impact and salary is reached. Here, I picked mean h-index ~ 11 (based on a google search) and correlation ~0.3 (because we wanted h-index to be only weakly important in this field)

eng_comp_grads <- eng_comp_grads |> 
  dplyr::mutate(h_index = eng_comp_grads$salary * 0.00005 + rnorm(1000, 5, 3))
mean_impact <- mean(eng_comp_grads$h_index)
correlation <- cor(eng_comp_grads$salary, eng_comp_grads$h_index)
hist(eng_comp_grads$h_index) #plot h-index values
```
